---
title: "DATA 606 Final Project - Relationship between Gun Ownership and Gun Violence"
author: "Bruno de Melo"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

### - Abstract

Objective is to analyze the relationship between deaths by firearms, either by self-harm or by physical violence, and gun ownership and gun availability in the United States of America.

Gun ownership in America is a very controversial topic, especially considering its constitutional status. This leads to a political debate that in most cases lacks factual basis based on data. Debate tends to heat up after mass shootings but it is also noteworthy that more than half of gun deaths are from suicides.

The USA possesses the highest level of gun ownership among the developed nations, around 1.2 per resident. Death by firearms when counted as a single category - which includes suicide, accidents and homicides -  kills about 40,000 American a year. This number is comparable to death by vehicle accident (39,000) or pancreatic cancer (45,000).

In light of gun availability that varies by state and by year, analysis focused on answering these questions:

- Is there a relationship between the number of deaths by firearm with the amount of firearms in a state? Does it change over time?

- Is there a relationship between the number of deaths by firearm with ownership restriction such as permit requirement and background check?

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(tidyverse)
library(leaflet)
library(ggpubr)
library(dotwhisker)
library(MASS)
```



### Part 1 - Introduction



Objective is to analyze the relationship between gun ownership and gun violence, either by self-harm or by physical violence. 




### Part 2 - Data


Two datasets are used:
1. Number of deaths by firearm (self-harm or by physical violence) by state.
2. State-Level Firearm Ownership by state.

None of the databases are directly accessible but rather require either a query to be built (dataset 1) or need to be downloaded in **zip** format (dataset 2).

Both datasets are include in the my github link: https://github.com/bsvmelo/DATA_606

Dataset 1: Number of deaths by firearm (self-harm or by physical violence) by state. It is available via a query that has to be built via a website, source is here: http://ghdx.healthdata.org/gbd-results-tool?params=gbd-api-2019-permalink/e89460e52883bcb25c4e9cdacb72cc09

Data was obtained on Global Health Data Exchange (GHDx), which is maintaining by the University of Washington. The Institute for Health Metrics and Evaluation (IHME) is an independent global health research center at the University of Washington. The Global Health Data Exchange (GHDx) is a data catalog created and supported by IHME.


Dataset 2: State-Level Firearm Ownership by state. Source is provided here https://www.rand.org/pubs/tools/TL354.html and a zipped Excel file is available for downloading.

Data was obtained on The RAND Corporation website, RAND is a research organization that develops solutions to public policy challenges to help make communities throughout the world safer and more secure, healthier and more prosperous. RAND is nonprofit, nonpartisan, and committed to the public interest.

Data summary of these two datasets is below.

```{r data load, echo=TRUE, results='hold', warning=FALSE, message=FALSE}
# load data

#Dataset1: csv file
ds1_loc<-"https://raw.githubusercontent.com/bsvmelo/DATA_606/main/IHME-GBD_2019_DATA-f5fee567-1.csv"
ds1<-data.frame(read.csv(ds1_loc))
#Dataset2: xlsx file
#https://github.com/bsvmelo/DATA_606/blob/main/TL-354-State-Level%20Estimates%20of%20Household%20Firearm%20Ownership.xlsx
#Dataset2: csv file
ds2_loc<-"https://raw.githubusercontent.com/bsvmelo/DATA_606/main/State-Level_Estimates_of_Household_Firearm_Ownership.csv"
ds2<-data.frame(read.csv(ds2_loc))
#check states are similar in two files
# states_1<-ds1 %>%
# group_by(location_name)  %>%
# summarise ()
# #nrow(states_1)
# states_2<-ds2  %>%
# group_by(STATE)  %>%
# summarise ()
# #nrow(states_2)
# #Number of states are not the same, but this will not pose a problem given that ds1 has one more states than ds2.

summary(ds1)
summary(ds2)

```

These two datasets are merged into one dataset - code in the chunk below. It contains 2,700 observations across 10 variables, as described below:

`year`: Describes Year (filtered to show common date between `ds1` and `ds2` datasets).
`State`: Describes State.
`cause_id`: Cause of death index.
`Cause_death`: Describes cause of death with two factors: Physical violence by firearm or Self-harm by firearm.
`No_Deaths`: Describes number of deaths.
`Ownership`: Describes percentage of gun ownership.
`Universal`: Describes gun restriction in relation required universal background check.
`permit`: Describes gun restriction in relation required permit to carry gun.
`abb`: Describes state abbreviation.
`region`: Describes State region.


```{r data prep 1}

#Data set Number of deaths

deaths<-ds1 %>%
filter(year <= 2016, measure_name == "Deaths", metric_id==1, location_name!="District of Columbia") %>%
select(year, location_name,cause_id,cause_name,val)

deaths_total<-ds1 %>%
filter(measure_name == "Deaths", metric_id==1, location_name!="District of Columbia") %>%
select(year, location_name,cause_id,cause_name,val)

#Data set Gun ownership
n_guns<-ds2 %>%
  filter(Year >= 1990) %>%
  select(Year, STATE,HFR,universl,permit) 

n_guns_total<-ds2 %>%
  select(Year, STATE,HFR,universl,permit)

#Merging
df<-left_join(deaths, n_guns, by = c("year" = "Year", "location_name"="STATE"))
colnames(df)<-c("Year","State","cause_id","Cause_Death","No_Deaths","Ownership","Universal","permit")

df$cause_id<-as.factor(df$cause_id)
df$State<-as.factor(df$State)
df$abb<-state.abb[df$State]
df$region<-state.region[df$State]

summary(df)
```


### Part 3 - Exploratory data analysis

As a context setting to the analysis, picture below depicts the ownership rate across states in 2016. We observe high ownership rates in the North Central and South regions, while lower rates are observed in the West and in the Northeast.

```{r}
states <- map_data("state")
df_2016<-df%>%filter(!is.na(Ownership), Year==2016)%>%
select(State,Ownership)%>%
mutate(Ownership = round(Ownership*100,1))
df_2016$State <- tolower(df_2016$State)

ggplot(df_2016, aes(map_id = State)) + geom_map(aes(fill = Ownership), map = states) +
expand_limits(x = states$long, y = states$lat)  + scale_fill_gradient(name = "Ownership-%", low = "white", high = "red") +
labs(title = "US Gun Ownership by States, 2016", x = "longitude", y = "latitude",
caption = "source: https://www.rand.org/pubs/tools/TL354.html") +
coord_fixed(1.3) +
theme(panel.background = element_blank())


```

Over time, ownership rate has been following in average and across the states, as shown below. Red line indicates the national average.


```{r}
n_guns_total%>%group_by(Year)%>%summarise(mean=mean(HFR)*100)%>%
          ggplot(mapping=aes(x=Year,y=mean)) +  geom_line(colour="red", size=2) +
        geom_line(n_guns_total,mapping=aes(x=Year,y=HFR*100, group=STATE),colour="gray80") +
        labs(title = "Gun Ownership by States since 1980, in % of population", x = "Year", y = "% Ownership Rate")
```


In terms of aggregated number of deaths over time of all states, chart below depicts the historical trend since 1990. Dataset contains information whether death was caused by self-harm (suicide) or by physical violence (variable `Cause_Death`).

Three conclusions can be drawn:
1: Overall trend indicates increasing deaths, especially after 2000.
2: Proportion of self-harm has sharply increased since 2005.
3: Proportion of non-self-harm decreased substantially from 1990 to 2000.

```{r}
a<-deaths_total%>%group_by(year)%>%summarise(sum1=sum(val))
b<-deaths_total%>%group_by(year,cause_name)%>%summarise(sum2=sum(val))



p1<-ggplot(a,mapping=aes(x=year,y=sum1)) + geom_line() + labs(title = "Number of Death by firearms since 1990", x = "Year", y="No of deaths")
p2<-ggplot(b)+geom_line(b,mapping=aes(x=year,y=sum2, color=cause_name))+guides(color=guide_legend("Cause of Death")) + labs(title = "", x = "Year", y="")
ggpubr::ggarrange(p1, p2, ncol = 2, nrow = 1)

```


Descriptive statistics on ownership data, shows that the national average since 1990 displays bi-modal distribution with distinct group of states with high and low rate of ownership. National mean is 42% with a standard deviation of 4.5%.

```{r}
rate_nat<-df%>%group_by(Year)%>%summarise(mean=mean(Ownership)*100)
ggplot(rate_nat,mapping=aes(x=mean))+geom_histogram(bins=15)
summary(rate_nat)
```

Boxplot per region is displayed below. Northeast region has the lowest median over time while South region has the highest.

```{r}
ggplot(df,aes(x=region, y=Ownership,group=region))+geom_boxplot()

```

Boxplot per year is displayed below. As evidenced in the aggregate chart above, median ownership has decreased overtime.

```{r}
ggplot(df,aes(x=Year, y=Ownership,group=Year))+geom_boxplot()

```

Descriptive statistics on number of deaths, shows that the national average since 1990 shows a quasi normal distribution. National mean is 34k with a standard deviation of 2,175.

```{r}
death_nat<-df%>%group_by(Year)%>%summarise(mean=mean(No_Deaths)*100)
ggplot(death_nat,mapping=aes(x=mean))+geom_histogram(bins=15)
summary(death_nat)
```

Boxplot per region is also displayed below. Lower median value is observed in the Northeast while the higher is observed in the South.

```{r}
ggplot(df,aes(x=region, y=No_Deaths,group=region))+geom_boxplot()

```

Boxplot per year is displayed below. As evidenced in the aggregate chart above, median number of deaths has increased overtime.

```{r}
ggplot(df,aes(x=Year, y=No_Deaths,group=Year))+geom_boxplot()

```


Below we check how law restrictions have changed over time by state. Two variables are associated with law restrictions:
`Universal`: Describes gun restriction in relation required universal background check.
`permit`: Describes gun restriction in relation required permit to carry gun.

These are boolean variables in that a $0$ means no restriction and a $1$ means restriction. Table below shows whether states are becoming more restrictive (changing from 1 to 0) or less restrictive (changing from 1 to 0) or have not change their laws over time.


\begin{table}[]
\begin{tabular}{llll}
\multicolumn{4}{l}{Universal Background Check}                       \\
\multicolumn{2}{l}{No Restriction} & \multicolumn{2}{l}{Restriction} \\
No Change     & Less Restrictive   & No Change    & More Restrictive \\
All others    & Missouri           & Hawaii       & California       \\
              & Tennessee          & Iowa         & Colorado         \\
              &                    & Michigan     & Connecticut      \\
              &                    & New York     & Delaware         \\
              &                    & Rhode Island & Maryland         \\
              &                    &              & Oregon           \\
              &                    &              & Pennsylvania     \\
              &                    &              & Washington      
\end{tabular}
\end{table}

\begin{table}[]
\begin{tabular}{llll}
\multicolumn{4}{l}{Permit to carry gun}                               \\
\multicolumn{2}{l}{No Restriction} & \multicolumn{2}{l}{Restriction}  \\
No Change     & Less Restrictive   & No Change     & More Restrictive \\
All others    & Missouri           & Hawaii        & California       \\
              &                    & Illinois      & Connecticut      \\
              &                    & Iowa          & Maryland         \\
              &                    & Massachusetts & Nebraska         \\
              &                    & Michigan      & North Carolina   \\
              &                    & New Jersey    &                  \\
              &                    & New York      &                  \\
              &                    & Rhode Island  &                 
\end{tabular}
\end{table}

#### Relationship between gun ownership rate, number of deaths and restriction laws

in this section, the relationship between gun ownership rate, number of deaths and restriction laws are visually explored.

Starting with ownership vs restriction laws by state as of 2016. Restriction law here refers to requirement of having a permit to carry a gun. Chart below is jittered for ease of visualization, hence the duplication. In general, the more restrictive the laws the lower is the ownership rate.

```{r}
set.seed(123)

df%>%filter(!is.na(Ownership), Year==2016)%>%
select(abb,Ownership, No_Deaths, permit, Universal,region)%>%
mutate(Ownership = round(Ownership*100,1))%>%
ggplot(aes(jitter(as.numeric(as.factor(permit)),0.5),Ownership,group=abb,colour =abb, label=abb))+geom_point(size=2, alpha=0)+geom_text(aes(label=abb),hjust=0, vjust=0.1, size=3)+ theme(legend.position = "none") +labs(title = "Ownership rate vs Restriction Laws", x = "Restriction Law", y="% Ownership Rate") + annotate("text", x = 1.1, y = 25, label = "Less restrictive", colour="Green")+ annotate("text", x = 1.9, y = 50, label = "More restrictive",colour="Orange")+ theme(axis.ticks = element_blank(),axis.text.x=element_blank())

```


Next, the relationship between number of deaths and ownership. Apart from the three outliers, CA, FL and TX, chart indicates an indirect relationship between number of deaths and ownership rate.

```{r}
set.seed(123)

df%>%filter(!is.na(Ownership), Year==2016)%>%
group_by(abb)%>%
mutate(Ownership = round(Ownership*100,1))%>%
summarize(total=sum(No_Deaths),Ownership=mean(Ownership))%>%
ggplot(aes(x=Ownership,y=total,group=abb,colour =abb, label=abb))+geom_point(size=2, alpha=0)+geom_text(aes(label=abb),hjust=0, vjust=0.1, size=3)+ theme(legend.position = "none") +labs(title = "Number of Deaths vs Ownership rate", x = "% Ownership Rate", y="Number of Deaths") + annotate("text", x = 20, y = 2000, label = "More restrictive", colour="Green")+ annotate("text", x = 50, y = 2000, label = "Less restrictive",colour="Orange")

```


### Part 4 - Inference

As seen in the previous section, four variables will be analysed through regression analysis in order to determine whether they are statistically related to the number of deaths: `ownership`, `permit`, `universal`, and `year`.

Dataset is in fact a collection of 50 time series (one per state) of ownership rate data and restriction laws. Given the time dimension aspect, panel data or fixed effect regression analysis technique will utilized. This was not covered in class but it can be run using the function `lm` that we are familiar with. Fixed-effects regression models are models that assume a non-hierarchical data structure, i.e. data where data points are not nested or grouped in higher order categories (source: https://slcladal.github.io/regression.html) like the `States` variable present in the dataset.
Fixed effects analysis can be viewed as one that controls for systematic differences across clusters, just like controlling for a categorical variable.

General equation is of the form:

\[

f_{(x)} = \alpha + \beta_{1}x_{ij} + \beta_{2}x_{ij} + \dots + \beta_{n}x_{ij} + \gamma_j + \varepsilon_{ij}

\]

where *i* indicates `year`, *j* indicates `state`, and \gamma_j indicates the *state fixed effect*. In practical terms, it indicates that we put in dummy variables for the states.

Analysis will be conducted first on the aggregate number of deaths and in sequence considering the two different causes of death: suicide and violence by gun.

#### Aggregate Analysis

Regression equation is of the form:

\[

\widehat{DeathsTotal} = \beta_0 + \beta_1\times (Ownership_{ij}) + \beta_2\times (Permit_{ij}) + \beta_3\times (Universal_{ij}) + \beta_4\times (Year_{ij}) + \gamma_j + \varepsilon_{ij}

\]

where \gamma_j the *state fixed effect*.

Results are presented in the table below; results for the 'State' variable are suppressed for better interpretation - available in the Appendix.

```{r}
#aggregating sum of deaths
df_agg<-df%>%filter(!is.na(Ownership))%>%
          group_by(State, Year,Ownership,permit,Universal)%>%
          summarise(No_Deaths = sum(No_Deaths))%>%
          mutate(rate= Ownership*100)

d_total<-lm(No_Deaths ~ rate + (permit) + (Universal) + Year + State  -1, df_agg)

m4_df <- coef(summary(d_total)) %>% 
  data.frame() %>% 
  tibble::rownames_to_column("term") %>%
  rename(estimate = Estimate, std.error = Std..Error)
(d_total_results<-m4_df[m4_df$term=='rate' | m4_df$term=='permit' | m4_df$term=='Universal' | m4_df$term=='Year',  ])

```



Resulting equation is below, and has an adjusted R-squared of 0.98:
\[

\widehat{DeathsTotal} = 7.04\times (Ownership) -150.5\times (Permit) + 3.62\times (Year) + \gamma_j

\]

It means that for a 1% increase in gun ownership rate, number of deaths increases by 7 people. Sign of `permit` variable is negative, meaning that states with a restriction on carrying gun with a permit see a decrease in the number of deaths of 150 people. The `Universal` variable is not significant, while `Year` variable is significant and its meaning suggests that the number of deaths increases over time, as highlighted in a previous chart. Result on all states are significant.

Diagnostic plots are show below. Residuals are slightly not well behaved, showing a trend in the first chart, Residuals vs Fitted. The normal Q-Q plot suggests that the residual are not normal, maybe due to outliers.

```{r}
# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page
plot(d_total)
```

stepwise selection (forward, backward, both) is performed using the `stepAIC` function from the *MASS* package. Function performs stepwise model selection by exact AIC. Final model suggested is the same as found above, without the `universal` variable.

```{r}
step <- stepAIC(d_total, direction="both")
step$anova # display results
```




#### Cause of Death Analysis

Considering that the categorical variable `cause_deaths` has two levels, two regression analysis will be run, following the same methodology as described above.

*Deaths by suicide*

Regression equation is of the form:

\[

\widehat{Deaths by Suicide} = \beta_0 + \beta_1\times (Ownership_{ij}) + \beta_2\times (Permit_{ij}) + \beta_3\times (Universal_{ij}) + \beta_4\times (Year_{ij}) + \gamma_j + \varepsilon_{ij}

\]

Results are presented in the table below; results for the 'State' variable are suppressed for better interpretation - available in the Appendix.

```{r}
#cause of death 721
df_721<-df%>%filter(!is.na(Ownership), cause_id==721)%>%
group_by(State, Year,Ownership,permit,Universal)%>%
          mutate(rate= Ownership*100)


d_721<-lm(No_Deaths ~ rate + (permit) + (Universal) + Year + State  -1, df_721)

m5_df <- coef(summary(d_721)) %>% 
  data.frame() %>% 
  tibble::rownames_to_column("term") %>%
  rename(estimate = Estimate, std.error = Std..Error)
(d_721_results<-m5_df[m5_df$term=='rate' | m5_df$term=='permit' | m5_df$term=='Universal' | m5_df$term=='Year',  ])

```



Resulting equation is below, and has an adjusted R-squared of 0.99:
\[

\widehat{DeathsTotal} = 2.3\times (Ownership) -48.8\times (Permit) + 3.62\times (Year) + \gamma_j

\]

Result means that for a 1% increase in gun ownership rate, number of suicides increase by 2 people. Sign of `permit` variable is negative, meaning that states with a restriction on carrying gun with a permit see a decrease in the number of suicides by 49 people. The `Universal` variable is not significant, while `Year` variable is significant and its meaning suggests that the number of suicides increases over time, as highlighted in a previous chart. Result on all states are significant.

Diagnostic plots are show below. Residuals are slightly not well behaved, showing a trend in the first chart, Residuals vs Fitted. The normal Q-Q plot suggests that the residual are not normal, maybe due to outliers.

```{r}
# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page
plot(d_721)
```

Stepwise selection (forward, backward, both) is performed using the `stepAIC` function from the *MASS* package. Final model suggested is the same as found above, without the `universal` variable.

```{r}
step <- stepAIC(d_721, direction="both")
step$anova # display results
```

*Deaths by violence*

Regression equation is of the form:

\[

\widehat{Deaths by Violence} = \beta_0 + \beta_1\times (Ownership_{ij}) + \beta_2\times (Permit_{ij}) + \beta_3\times (Universal_{ij}) + \beta_4\times (Year_{ij}) + \gamma_j + \varepsilon_{ij}

\]

Results are presented in the table below; results for the 'State' variable are suppressed for better interpretation - available in the Appendix.

```{r}
#cause of death 721
df_725<-df%>%filter(!is.na(Ownership), cause_id==725)%>%
group_by(State, Year,Ownership,permit,Universal)%>%
          mutate(rate= Ownership*100)


d_725<-lm(No_Deaths ~ rate + (permit) + (Universal) + Year + State  -1, df_725)

m6_df <- coef(summary(d_725)) %>% 
  data.frame() %>% 
  tibble::rownames_to_column("term") %>%
  rename(estimate = Estimate, std.error = Std..Error)
d_725_results<-m6_df[m6_df$term=='rate' | m6_df$term=='permit' | m6_df$term=='Universal' | m6_df$term=='Year' ,  ]

(d_725_results_st<-m6_df[m6_df$Pr...t..<=0.1,  ])


```



Resulting equation is below, and has an adjusted R-squared of 0.95:
\[

\widehat{DeathsTotal} = 4.7\times (Ownership) -101.6\times (Permit) + 36\times (Universal) + \gamma_j

\]

Result means that for a 1% increase in gun ownership rate, number of deaths increase by 5 people. Sign of `permit` variable is negative, meaning that states with a restriction on carrying gun with a permit see a decrease in the number of deaths by 102 people. The `Universal` variable is significant at th 95% level, and indicates states that require a background check see an increase in deaths by 37 people, which seems to be a contracting result. Variable `Year` is not significant.

Only 10 states show statistical significance at least the 10% level, differently from the previous two regressions. Given the positive signal, this means that number of deaths by gun violence increases over time in these states.


Diagnostic plots are show below. Residuals are slightly better behaved, showing no trend in the first chart, Residuals vs Fitted. The normal Q-Q plot suggests that the residual are not normal, maybe due to outliers.

```{r}
# diagnostic plots
layout(matrix(c(1,2,3,4),2,2)) # optional 4 graphs/page
plot(d_725)
```

Stepwise selection (forward, backward, both) is performed using the `stepAIC` function from the *MASS* package. Final model suggested is the same as found above, without the `universal` variable.

```{r}
step <- stepAIC(d_725, direction="both")
step$anova # display results
```


#####Putting all together


```{r}

d_total_results<-d_total_results%>% mutate(model = "Aggregate",p_value=ifelse(Pr...t..<0.001,0,round(Pr...t..,2)))

d_721_results<-d_721_results%>% mutate(model = "Suicide",p_value=ifelse(Pr...t..<0.001,0,round(Pr...t..,2)))
d_725_results<-d_725_results%>% mutate(model = "Violence",p_value=ifelse(Pr...t..<0.001,0,round(Pr...t..,2)))

all_models <- rbind(d_total_results,d_721_results,d_725_results)
all_models%>%filter(p_value<0.05)%>%
dwplot() + theme_bw() + xlab("Coefficient Estimate") + ylab("") +
    ggtitle("Gun Violence Regression Models", subtitle = "with significant coefficents")


```




We will regression analysis to answer the questions

- Is there a relationship between the number of deaths by firearm with the amount of firearms in a state? Does it change over time?

- Is there a relationship between the number of deaths by firearm with ownership restriction such as permit requirement and background check?









### Part 5 - Conclusion


### References


### Appendix (optional)
